### 2014-04-03 paper/designspace.tex
working through comments, now time for george to read

\section{The Design Space of Pipes}
\tovi{clean up writing}By using different openings, designs, and topologies of pipes, and inserting different media into the completed pipes, inputs and outputs can be located at arbitrary points on the surface of a 3D print.  We describe the space below.

@@

Pipe network topologies enable different interactions.  Splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).  Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out. \tovi{give simple examples of where this might be useful}

@@

Incompressible fluids (``liquids'') perform similarly to gases.  Additionally, using driable conductive liquids like copper paint, pipes can become custom wires. \tovi{custom water jet devices (uist 2013 competition)}

>>>>

\section{The Design Space of Pipes}
The design space of pipes has four dimensions: openings, design focus, topologies, and inserted media.  We describe the space below.

@@

Pipe network topologies enable different interactions.  Splitting or mixing pipes offer flexibility in output, allowing a single source to have multiple outputs or multiple sources to combine (see Figure \ref{fig:pipespace}).  Star and tree topologies extend the splitting and mixing primitives, encompassing multiple-in and multiple-out in contrast to single-in/single-out; we used star topologies in our toys \ref{fig:toys}.

@@

Incompressible fluids (``liquids'') perform similarly to gases for display or haptics.  Additionally, using driable conductive liquids like copper paint, pipes can become custom wires.

### 2014-04-03 paper/discussion.tex
first set of glamour shots, paper work, redo of figures on new mm background, redo some figures due to new modeling
working through comments, now time for george to read

The promise of 3D printing is that it allows for the manufacture of geometries that are impossible to create using traditional techniques like injection molding.  Pipes leverage that promise: designing the internal structure of an object was not previously feasible.  As printers at the hobbyist and corporate levels become more sophisticated and able to fabricate intricate geometries of ever more materials, we believe that our tool will still assist in design tasks.  Even if printers are capable of laying down conductive, optically transparent, or luminescent materials in-place, the process of designing their paths is still crucial.

By crossing pipe types, topologies, and inserted media, we create a space of possible inputs and outputs that can be pipe-mediated.  We arrange our I/O discussion around the modes of input or output (the five senses, touch/tapping/pressure/etc.).  We also suggest new points in the design space of pipe-mediated interaction that remain unexplored.

@@

\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.  Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke (as in the Aireal \cite{Sodhi-aireal} demo) also offers visual feedback.  Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.  Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.  Printed optics \cite{Willis-printedoptics} explored visual output at the ends of pipes, and  EL wire can be threaded through custom paths to create neon signs, as in our neon robot (Figure \ref{fig:neon}).

@@

\emph{Haptic} outputs are possible through compressible and incompressible fluids.  Fluids in pipes capped with flexible membranes can actuate particulates and create programmable pliability as in Jamming User Interfaces \cite{Follmer-jamming}.  Additionally, use of semi-closed pipes with rubberlike material as the caps allows for gases to actuate surface features, like the back of our breathing rabbit (see Figure \ref{fig:breathe}).  Free-air haptic feedback as in Aireal \cite{Sodhi-aireal} and volumetric interactions as in Volflex \cite{Iwata-volflex} can also be pipe-mediated.

@@

\emph{Touch} sensing can be enabled through pipes filled with conductive media.  Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.  Using conductive paint and Swept Frequency Capacitive Sensing (SFCS \cite{Sato-touche}), an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface, like our toys (see Figure \cite{fig:toys}).  Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.  Using a technique like Fiberio \cite{Holz-fiberio}, users could even be identified on touches.

@@

Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.  Components can be recessed into a print's surface, with their leads implanted into open pipes.  By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place (see Figure \ref{fig:radio}).

@@

We have also experimented with intentional encoding.  Semi-closed pipes can function as resonance chambers.  An object's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.  Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.  Intentional encoding in the case of capacitance is significantly more difficult, and we leave it to future work.  Encoding of weight is trivial by creating fully-enclosed chambers inside an object.

@@

>>>>

The promise of 3D printing is that it allows for the manufacture of geometries that are impossible to create using traditional techniques like injection molding.  Pipes leverage that promise: they require a previously unmanufacturable complexity of interior geometry.  Printers additionally continue to evolve in their capabilities to lay down more materials.  As they become able to lay down conductive, optically clear, or luminescent materials, our tools and techniques will enable even more complex internal designs.

By crossing pipe types, topologies, and inserted media, many pipe-mediated interactions emerge. We demonstrated a few such possibilities by designing and fabricating example objects with the help of our pipe routing design tool. However, there are many additional possibilities for leveraging sensing and actuation techniques from the HCI literature and extending their use in 3D-printed objects. We briefly survey such possibilities for output and input.

@@

\emph{Visual} outputs can be mediated by gases, liquids, solid, particulates, or threadables.  Pneumatic actuation (as in PneUIs \cite{Yao-pneui} or Harrison's latex buttons \cite{Harrison-buttons}) provides output, but gases carrying smoke also offer visual feedback.  Splitting and mixing of colored liquids can be used as a display tactic; additionally mechanical motion can be fluid-mediated.  Liquids can also be mixed in specific concentrations by servo-controlled gate structures as in Figure \ref{fig:direct}.

@@

\emph{Haptic} outputs are possible through compressible and incompressible fluids.  Fluids in pipes capped with flexible membranes can actuate particulates and create programmable pliability as in Jamming User Interfaces \cite{Follmer-jamming}.

@@

\emph{Touch} sensing can be enabled through pipes filled with conductive media.  Traditional wires (threadable) may be used, but conductive paint (liquid) may be more flexible.  Using conductive paint and SFCS \cite{Sato-touche}, an interior star pipe topology can enable single-wire touch- and grasp-sensing at any set of points on a printed object's surface, like our toys (see Figure \cite{fig:toys}).  Simple capacitance measurements are possible using the same media, and custom-designed sensors like those created in Savage, et al.,'s Midas system are also possible \cite{Savage-midas}.  Using a technique like Fiberio \cite{Holz-fiberio}, users could even be identified on touches.

@@

Sensing via traditional electronic components (e.g., potentiometers, alcohol gas sensors) can be accomplished by conductive inserted media.  Components can be recessed into a print's surface, with their leads implanted into open pipes.  By inserting liquid copper paint instead of threading traditional wires, all components in a model can share a ground line, and the paint's drying process obviates the use of solder or glue to affix the components in place.

@@

We have experimented with intentional encoding.  Semi-closed pipes can function as resonance chambers.  A pipe's resonant frequency can be measured by attaching a speaker and microphone to it, sweeping frequencies with the speaker, and performing a Fourier transform on the resultant signal from the microphone.  Peaks in the transformed data correspond to stronger returned impulses: the resonant frequencies of the object.  Intentional encoding in the case of capacitance is significantly more difficult, and we leave it to future work.  Encoding of weight is trivial by creating fully-enclosed chambers inside an object.

@@

\bjoern{We might pull back on the scope of this section a bit and think about which directions here are most closely related to pipes and most likely successful;  and which are more tangential or speculative - for example, Aireal is about feedback in free space and we are all about designing tangible objects; so maybe that's a candidate for cutting.}

\valkyrie{I cut a few things, and removed references to techniques we have already shown to work (as in the spirit of the intro).  LMK if this is better.}

### 2014-04-03 paper/introduction.tex
working through comments, now time for george to read

\section{Introduction}
\tovi{Typical intro format: first pragraph should not introduce your own work, but summarize the problem and motivation. Second paragraph should summarize prior art and its limations. After that you can describe your work and contributions. I've made some revisions to the strucutre...}
Makers, as well as professional designers, leverage 3D printers as tools for design work.  A wide array of objects, ranging from bicycle helmets to video game controllers, are now prototyped or even manufactured using these machines.  However, most devices fabricated by 3D printers are passive, prototyping the form, but not the interactivity, of an object.

Due to the growing popularity of 3D printing and Maker Communities [ref?], recent work has looked at the issue of rapidly prototyping interactive objects. Existing approaches include ... but their limitations are ...

We propose a novel technique involving the removal of material from the interior of 3D models, prior to printing, to create pipes and other cavities.  These pipes can be filled, post-print, by a variety of media that enable input, display, and tactile feedback.  This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process [ref]. While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to rapidly prototype a diverse range of interactive objects on commonly available 3D printers (Figure \ref{fig:teaser}).\tovi{I guess we need to be careful here since not all of our examples are on doen with commonly available 3D printers}

We describe a new design space of pipes and hollow chambers for the prupose of interaction design, where variables include openings, topologies, and inserted media. By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized. For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.\tovi{at some point explain why this is better than running wires on the exterior (looks better, for one)}

\tovi{I would remove this entire paragraph}
Pipes allow makers to build with and upon existing input/output techniques, often integrating these methods into objects they were not originally designed for.   For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the brain.  After creating pipes to those locations and printing her object, she can fill the pipes with conductive paint.  For sensing, she need only connect a single wire to a shared out of all the tubes, and can distinguish them via Swept Frequency Capacitve Sensing \cite{Sato-touche}.  In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several sensing locations.  Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on an print's surface.  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.

\tovi{I would remove this entire paragraph}
Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.  Electricity is run through the near-vacuum, creating light.  If our maker wants to build such a sign, she can substitute EL wire for the evacuated glass: she need only design a path for it to be fed through and then create pipes along that path.

>>>>

\section{Introduction}
Makers, as well as professional designers, leverage 3D printers as tools for design work.  A wide array of objects, ranging from bicycle helmets to video game controllers, are now prototyped or even manufactured using these machines.  However, most devices fabricated by 3D printers are passive, prototyping the form, but not the interactivity, of an object.

Due to the growing popularity of 3D printing and maker communities like Thingiverse\footnote{http://thingiverse.com}, recent work has looked at the issue of rapidly prototyping interactive objects. Existing approaches include Printed Optics \cite{Willis-printedoptics}, which uses optically clear material for light redirection, and Sauron \cite{Savage-sauron}, which is based on computer vision, but they are limited to visible interactions only; they do not give designers freedom to integrate haptic or acoustic inputs or outputs.

We propose a novel technique involving the removal of material from the interior of 3D models, prior to printing, to create pipes and other cavities.  These pipes can be filled, post-print, by a variety of media that enable input, display, and tactile feedback.  This {\em subtractive} approach is complementary to {\em additive} approaches that inject custom print materials, such as conductors, during the actual printing process [ref]. While our approach requires some manual assembly after a print completes, it gives makers the opprotunity to rapidly prototype a diverse range of interactive objects on 3D printers, including hobbyist machines (Figure \ref{fig:teaser}).

We describe a new design space of pipes and hollow chambers for the prupose of interaction design, where variables include openings, topologies, and inserted media. By exploring this design space new opportunities for adding interactivity to 3D printed objects can be realized. For example, copper material can fill the channels, to allow for standard electronic components to be easily integrated after the printout; electroluminescent (EL) wire can be threaded through to create computer-controlled visual output; or air can be pumped through to control haptic responses.  In comparison to traditional exterior electronics integration, our process allows makers to preserve aesthetics by hiding wires inside the object; additionally it offers more freedom in selecting precise surface locations for, e.g., air mediated tactile output.

\tovi{I would remove this entire paragraph} \valkyrie{I think we at least need to preserve the sentiment from the end, that we don't have any novel I/O techniques but that pipes help extend existing techniques in useful ways}
Pipes allow makers to build with and upon existing input/output techniques, often integrating these methods into objects they were not originally designed for.   For example, to create a touch-sensitive toy as in Figure \ref{fig:teaser}, a maker can select several locations of interest on the brain.  After creating pipes to those locations and printing her object, she can fill the pipes with conductive paint.  For sensing, she need only connect a single wire to a shared out of all the tubes, and can distinguish them via Swept Frequency Capacitve Sensing \cite{Sato-touche}.  In this way, she has only a single active element (the microcontroller setup), but her otherwise-inert 3D printed object (the brain) has several sensing locations.  Other existing sensing and actuation approaches, such as FlyEye \cite{Wimmer-flyeye} and Jamming User Interfaces \cite{Follmer-jamming}, can also be enhanced by the use of pipes: makers utilizing such I/O strategies can locate input and output at arbitrary points on an print's surface.  We leverage these existing techniques for sensing and actuation while our work's novelty is in \emph{internal modeling} for 3D printing and the \emph{redirection} of I/O to arbitrary locations on a device.

\tovi{I would remove this entire paragraph}  \valkyrie{if we scrap the first one, sure.  George, what do you think?}
Pipes can also be used for sending media on a specified route through an object; for example, neon signs are created by evacuated glass tubes formed into a specific path.  Electricity is run through the near-vacuum, creating light.  If our maker wants to build such a sign, she can substitute EL wire for the evacuated glass: she need only design a path for it to be fed through and then create pipes along that path.

### 2014-04-03 paper/limitations.tex
working through comments, now time for george to read

Some media are difficult to insert in pipes, or may not be compatible with all types of pipes.  Fluids in particular provide challenges.  For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.  Any liquid kept in tubes printed by the Objet can cause the exterior of the part to discolor due to seepage.  Hobbyist machines do not guarantee watertightness of their prints, so leaking of water or air can occur.

>>>>

Some media are difficult to insert in pipes, or may not be compatible with all types of pipes.  Threadable materials in particular can provide challenges, as they can get stuck while being fed into pipes.  Fluids also  For example, the conductive paint we used in our designs can take multiple days to dry in tubes of sufficient length.  Any liquid kept in tubes printed by the Objet can cause the exterior of the part to discolor due to seepage.  Hobbyist machines do not guarantee watertightness of their prints, so leaking of water or air can occur.

As discussed, support material removal can also be a challenge.  Particularly on multi-jetting printers, printing support-free is not possible for most geometries, while for hobbyist-grade printers physically detaching printed support is not always trivial.

### 2014-04-03 paper/relatedwork.tex
working through comments, now time for george to read

\tovi{dont use citations as nouns}
Slyper, et al., in \cite{Slyper-pressure} 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure.  \tovi{clean up writing - finish telling the reader what they did or didnt do, then respond with what we do differently}This technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface; they had to place them in locations accessible for electronics insertion. \tovi{not sure this is relevant enough to cite here}Acoustic Barcodes explored integrating ridges into fabricated objects, whose acoustic signatures when scraped provided an ID code \cite{Harrison-acoustic}; while pipes can be used for identification and to create aural feedback, this is not our only goal.  Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.  While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.

@@

\subsubsection{Fabricating Channels}

Other research has investigated the integration of traditional electronics into 3D prints.  Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.  Sarik, et al., \cite{Sarik-tracebrush} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto objects to create circuitry.  Injection of liquid metals for sensing in softer-than-skin electronics was explored by Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature}.  Navarette, et al., created a functioning GPS routed in true 3D (not multilayer boards' 2.5D), hand-built using a one-off routing.  Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.   \tovi{relation of this stuff to ours isn't so clear}

@@

There is prior work on modeling the interior of objects (as opposed to their exterior).  In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}.  Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.  Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.''  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.  All these contributions are orthogonal to ours, as our focus is on modeling pipes and hollow chambers for fabricating interactive objects.   \tovi{need better summary statement instead of ``all of this is orthogonal''}

>>>>

Slyper, et al., 3D printed flexible robot armatures capable of sensing interaction via changes in air pressure \cite{Slyper-pressure}.  They created all their models by hand, and were forced to place interaction locations in areas accessible to electronics assembly: this technique is complementary to ours, as we allow placing these interactive elements anywhere on an object's surface. Sauron investigated computer vision to track interactions with physical mechanisms \cite{Savage-sauron}.  While Savage, et al., used mirrors for redirection, we believe that the required line-of-sight operation for their computer vision techniques may be possible using printed optics or inserted optical fibers.

@@

\subsubsection{Fabrication + Electronics}

Other research has investigated the integration of traditional electronics into 3D prints.  Sells, et al., \cite{Sells-reprap}, describe using a heated syringe to deposit metal in printed 2D circuitboard channels on objects' surfaces.  Sarik, et al., \cite{Sarik-tracebrush} and the Optomec company \cite{optomec} describe \emph{spraying} conductive material onto printed objects to create circuitry.  Injection of liquid metals for sensing in softer-than-skin electronics was explored by Park, et al., \cite{Park-microchannels} and Majidi, et al., \cite{Majidi-curvature}.  Navarette, et al., created a functioning GPS routed in true 3D (not multilayer boards' 2.5D), hand-built using a one-off routing.  Our work differs from these because we offer \emph{algorithms and techniques} for \emph{internal}, \emph{true 3D} routing.   \tovi{relation of this stuff to ours isn't so clear}

@@

There is prior work on modeling the interior of objects (as opposed to their exterior).  Takayama, et al., built a system described in \cite{Takayama-fruit} which allowed users to create internal color maps of the interior of objects, allowing them to see digital ``slices'' of their ``oranges.''  Our work focuses on 3D printable geometry rather than rendering.  In Make it Stand, material is removed from the interior of objects with the goal of changing their center of gravity \cite{Prevost-makeitstand}, and Infrastructs created encoded interior cavities in 3D printed objects designed to be sensed with a terahertz imaging tool.  Modeling such chambers is one aspect of our work, however our focus is on interaction rather than balance or identification.  Bickel, et al., explored enforcing deformation behavior with material blending \cite{Bickel-deformation}, while B\"{a}cher, et al., built an algorithm for automatically placing 3D fabricatable joints into meshes \cite{Bacher-articulated}.  Sauron automatically modified the interior of existing mechanical objects to make them amenable to vision-based sensing \cite{Savage-sauron}.  All these contributions are orthogonal to ours, as our focus is on precisely \emph{removing} material from printed objects, rather than blending materials or modifying components.
